GMSE: an R package for generalised management strategy evaluation

### 2018-04-18 notebook/ms/SI3.Rmd
More SI edits

@Johnson2018 recently estimated key demographic parameters for the Taiga Bean Geese for the Central Management Unit that includes geese that breed in "Northern most Sweden, Northern Norway, Northern and Central Finland, and adjacent North-wester parts of Russia, wintering mostly in Southern Sweden and South-east Denmark" [@AEWA2016]. They estimated goose survival under ideal conditions to be ca 0.878; this can be interpreted in our model by setting mortality to `remove_pr = 1 - 0.878`. Similarly, @Johnson2018 estimated intrinsic population growth rate and carrying capacity to be 0.15 and 93870, respectively, so we set `lambda = 0.183` (for convenience of demonstration, the upper 90% confidence interval of their estimate) and `res_death_K = 93870`. The global abundance of Taiga Bean Geese in 2009 was ca 63000 [@Fox2010], and ca 35000 in the Central Management Unit [@AEWA2016], which we can take as a starting abundance for our simulations (`RESOURCE_ini = 35000`). And the International Single Species Action Plan has a target population size of ca 70000 in the Central Management Unit [@AEWA2016], which we can use as a management target (`manage_target = 70000`).

For simplicity, here we assess confict using only the `gmse` function to show how parameter values can be set to provide useful results. Our goal is not to provide a detailed case study of the Taiga Bean Geese, but rather to demonstrate how such a case study would be possible using the `gmse` function. Additionally, all of the analysis that follows would also be possibe using the more detailed, but also more flexible, `gmse_apply` function. Where available, we use estimated demographic parameter values from @Johnson2018 and @AEWA2016. Where GMSE parameter values are not available, we use reasonable values or GMSE defaults. To make model inferences for real case studies, we strongly recommend simulating across a range of parameter values when empirical estimates are unavailable, as social-ecological dynamics might be sensitive to these unknown values.

@@

Below, we first allow only the culling of geese by users and plot the dynamics of the social-ecological system from a single simulation. Next, we run the same simulation but allow users to scare geese from their landscape cells; we then use the model to make inferences regarding how scaring as a management option might affect goose population dynamics, agricultural production, and conservation conflict in the system. All simulations are run using the `gmse` function, but novice R users may prefer run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line. Alternatively, experienced R users may prefer simulate by looping time steps through `gmse_apply`, which allows more flexibility for incorporating custom submodels and dynamically adjusting parameter values. We emphasise that the simulations below are intended only to demonstrate one use of GMSE on a species of conservation interest, not to make recommendations for management of Taiga Bean Geese.

@@

The results of the above simulation are plotted in Figure SI3-1 below.

```{r, eval = TRUE, fig.height = 7, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit. This example includes 80 farmers whose objective is to maximise their agricultural output, and one manager whose objective is to keep geese at a target abundance, over 20 simulated time steps. Goose locations at the end of the simulation are shown in the upper left panel, while the upper right panel shows the same landscape broken down among the 80 farmers (upper 60% of the landscape in multiple colours), along with non-agricultural land (lower 40% of the landscape in blue). Actual goose abundance is shown in the middle left panel (black solid line), along with its estimate by the manager (blue solid line). Horizontal red and blue dotted lines show landscape-level goose carrying capacity enacted on adult mortality and the manager's target for goose abundance, respectively. The organge line shows the total percent of landscape cells (including non-farmed cells) yield, as decreased by geese. The middle right panel shows this yield for each farmer, and for the public land (lower line in blue). The lower left panel shows the cost of culling for farmers, as set by the manager, and the lower right panel shows the total number of culls attempted by farmers over time."}
plot_gmse_results(res = sim_1$resource, obs = sim_1$observation, 
                  land = sim_1$land, agents = sim_1$agents, paras = sim_1$paras, 
                  ACTION = sim_1$action, COST = sim_1$cost);
```

The above Figure SI3-1 shows the dynamics of goose abundance and agricultural yield, along with how managers react to change in abundance and farmers react to manager policy. In the case of the simulation above, goose population begins to decline, causing managers to set policy to increase the cost of culling. This higher cost of culling restricts farmers from culling on their land, leading to fewer total culling actions. In this case, the abundance of geese continues downward, while agricultural yield increases over time. We can investigate the conflict between management policy and farmers using the `plot_gmse_effort` function in GMSE.

@@

Black lines indicate how permissive a manager is toward a particular action on a scale of 0 to 100, while coloured lines indicate how much effort farmers expend on a given action. When black lines are far below coloured lines, we can (cautiously) interpret this as a conflict between management of the goose population and farmer's interest in agricultural production. These time periods represent instances in which managers are not at very permissive of a particular action (in this case culling), but farmers continue to expend effort to do the action anyway.

```{r}
sim_2 <- gmse(time_max = 30, scaring = TRUE, plotting = FALSE);
```

@@

The results are plotted below.

>>>>

@Johnson2018 recently estimated key demographic parameters for the Taiga Bean Geese for the Central Management Unit that includes geese that breed in "Northern most Sweden, Northern Norway, Northern and Central Finland, and adjacent North-wester parts of Russia, wintering mostly in Southern Sweden and South-east Denmark" [@AEWA2016]. They estimated goose survival under ideal conditions to be ca 0.878; this can be interpreted in our model by setting mortality to `remove_pr = 1 - 0.878`. Similarly, @Johnson2018 estimated intrinsic reproductive rate and carrying capacity to be 0.55 and 93870, respectively, so we set `lambda = 0.275` (for simplicity, we simply use half the mean reproductive rate; GMSE does not currently distinguish female and male individuals) and `res_death_K = 93870`. The global abundance of Taiga Bean Geese in 2009 was ca 63000 [@Fox2010], and ca 35000 in the Central Management Unit [@AEWA2016], which we can take as a starting abundance for our simulations (`RESOURCE_ini = 35000`). And the International Single Species Action Plan has a target population size of ca 70000 in the Central Management Unit [@AEWA2016], which we can use as a management target (`manage_target = 70000`).

Our goal is not to provide a detailed case study of the Taiga Bean Geese, but rather to demonstrate how such a case study would be possible in GMSE. For simplicity, here we assess confict using only the `gmse` function to show how parameter values can be set to provide useful results. Novice R users may prefer run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line. Alternatively, experienced R users may prefer simulate by looping time steps through `gmse_apply`, which allows more flexibility for incorporating custom submodels and dynamically adjusting parameter values. Where available, we use estimated demographic parameter values from @Johnson2018 and @AEWA2016. Where GMSE parameter values are not available, we use reasonable values or GMSE defaults. To make model inferences for real case studies, we strongly recommend simulating across a range of parameter values when empirical estimates are unavailable, as social-ecological dynamics might be sensitive to these unknown values.

@@

Below, we first allow only the culling of geese by users and plot the dynamics of the social-ecological system from a single simulation. Next, we run the same simulation but allow users to scare geese from their landscape cells; we then use the model to make inferences regarding how scaring as a management option might affect goose population dynamics, agricultural production, and conservation conflict in the system. We emphasise that the simulations below are intended only to demonstrate one use of GMSE on a species of conservation interest, not to make recommendations for management of Taiga Bean Geese.

@@

The results of the above simulation are plotted in Figure 1 below.

```{r, eval = TRUE, fig.height = 7, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit. This example includes 80 farmers whose objective is to maximise their agricultural output, and one manager whose objective is to keep geese at a target abundance, over 20 simulated time steps. Goose locations at the end of the simulation are shown in the upper left panel, while the upper right panel shows the same landscape broken down among the 80 farmers (upper 60% of the landscape in multiple colours), along with non-agricultural land (lower 40% of the landscape in blue). Actual goose abundance is shown in the middle left panel (black solid line), along with its estimate by the manager (blue solid line). The horizontal blue dotted line shows the manager's target for goose abundance. The organge line shows the total percent of landscape cells (including non-farmed cells) yield, as decreased by geese. The middle right panel shows this yield for each farmer, and for the public land (lower line in blue). The lower left panel shows the cost of culling for farmers, as set by the manager, and the lower right panel shows the total number of culls attempted by farmers over time."}
plot_gmse_results(res = sim_1$resource, obs = sim_1$observation, 
                  land = sim_1$land, agents = sim_1$agents, paras = sim_1$paras, 
                  ACTION = sim_1$action, COST = sim_1$cost);
```

Figure 1 shows the dynamics of goose abundance and agricultural yield, along with how managers react to change in abundance and farmers react to manager policy. In the case of the simulation above, managers quickly set a policy of high cost for culling, which leads to a rise in the goose population and a decrease in crop yield for farmers. After roughly 20 time steps, the goose population rises above the manager target, at which point the manager becomes more permissive of culling and the cost of culling for users therefore declines. In response, users begin to cull geese on their land, and the goose population begins to stabilise around the target of 7000 total geese. We can investigate the conflict between management policy and farmers more directly using the `plot_gmse_effort` function in GMSE.

@@

Black lines indicate how permissive a manager is toward a particular action on a scale of 0 to 100, while coloured lines indicate how much effort farmers expend on a given action. When black lines are far below coloured lines, we can (cautiously) interpret this as a conflict between management of the goose population and farmer's interest in agricultural production. These time periods represent instances in which the manager is not at very permissive of a particular action (in this case culling), but farmers continue to expend effort to do the action anyway. In the case of the above simulation of potential conservation conflict between farmers and goose conservation, conflict is highest before time step 20, where the manager is not permissive of culling because the population is below the manager's target. Once the goose population has increased above the manager's target, conflict decreases because the desired culling is permitted by managers to keep the population at a target abundance. It is worth noting that, despite conflict as we define it decreasing, agricultural damage is still relatively high after the target goose population is achieved (Figure 1). Hence, on a broader scale, conflict might persist around the appropriate target population size rather than what actions are permitted for farmers; currently, this potential aspect of conflict is not modelled, but future versions of GMSE may attempt to incorporate such additional complexity in conflict scenarios.

We can model the consequences for goose population dynamics, agricultural produciton, and conservation conflict when scaring is a policy option available to the manager. The code below is a simulation identical to the one just described, but with a scaring option included through `scaring = TRUE`.

```{r, echo = FALSE}
sim_2 <- gmse(time_max = 30, scaring = TRUE, plotting = FALSE);
```

@@

The results are plotted in Figure 3.
